---
title: "Measures of Location"
runningheader: "Measures of Location" # only for pdf output
subtitle: "Measuring Central Tendancy" # only for html output
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library('plyr')
library(epiDisplay)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

```{marginfigure}
**Sample Statistic**: 
A numerical value used as a summary measure for a sample such as the sample
mean, $\bar{x}$. 
```

```{marginfigure}
**Population Parameter**: 
A numerical value used as a summary measure for a population such as the 
population mean, $\mu$. 
```

In previous discussions, we introduced the difference between and population
and a sample. This distinction matters as we start our discussion of 
measures of location or measures of central tendency. Measures based 
on samples are called **sample statistics**, Measures based on a population
are called **population parameters**. 

# Arithmetic Mean

```{marginfigure}
**Sample Mean**: 
\Large
\begin{equation} 
    \bar{x} = \frac{\Sigma x_i}{n}
\end{equation}
where $\Sigma$ means the sum of, $x_i$ is the value of x for the ith observation
and $n$ is the number of observations in the sample data set.
```

Measures of location attempt to capture the most typical values found in a 
variable. The most important measure of location is the **arithmetic mean**, or 
the average value, of a variable. The mean is computed by summing the data 
values and dividing by the number of observations. 
Let's walk through the components of the equation for the sample mean.\

```{marginfigure}
**Population Mean**: 
\Large
\begin{equation} 
    \mu = \frac{\Sigma x_i}{N}
\end{equation}
where $\Sigma$ means the sum of, $x_i$ is the value of x for the ith observation
and $N$ is the number of observations in the population data set.
```

\
**$\bar{x}$**, pronounced *x bar*, is the symbol for the sample mean.\
\
**$\Sigma$** means sum the values. So $\Sigma x_i$ means sum the values 
from $x_i$ to $x_n$. $i$ is referred to as an index.\
\
**$n$** is the size of the sample or the number of observations in the sample.\
\
Let's calculate the arithmetic mean for the following sample:\
\
**46, 54, 42, 46, 32**\
\
In this example, $x_1$=46, $x_2$=54, $x_3$=42, $x_4$=46, and $x_5$=32. For
the sake of simplicity, we often lay out our calculations as follows.

## Step 1: Calculate the sum of the values

$_i$|$x_i$
-|-----
1| 46
2| 54
3| 42
4| 46
5| 32
-|-----
$\Sigma$|`r sum(46,54,42,46,32)`



**$\Sigma x_i$** = 46 + 54 + 42 + 46 + 32 = `r sum(46,54,42,46,32)`

## Step 2: Divide by the number of observations

The number of observations is 5.\
\
**$n$** = 5\
\
Now that we know the two components of the formula,
$\bar{x} = \frac{\Sigma x_i}{n}$, we only need to divide them to arrive
at our answer.\
\
**$\bar{x}$**= 220/5 = `r mean(c(46,54,42,46,32))`\
\
The sample arithmetic mean (**$\bar{x}$**) in this example is `r mean(c(46,54,42,46,32))`.\

# Median

The **median** is another measure of location frequently used. The median is the
middle value in a sorted data set. For an **odd number of observations**, the median
is the middle number. For example, given the sample data in our previous 
example, the median is:

## Step 1: Sort the Data

**46, 54, 42, 46, 32** becomes ordered data set **32, 42, 46, 46, 54**

$_i$|$x_i$
-|-----
1| 32
2| 42
3| 46
4| 46
5| 54

## Step 2: Find the Middle Value

The middle value (the 3rd value of 5) is 46. The median is `r median(c(46,54,42,46,32))`. 
You will notice that the median equals the mean calculated
in our previous example for these data.

## Calculating the Median with an Even Number of Observations

When calculating the median with an **even number of observations**, first sort
the data. Given the following sorted data set, we find the two middle values
and calculate their mean to find the median.\
\
**32, 42, 44, 46, 46, 54**\
\
(44 + 46)/2 = 45 is the median in this example.

# Comparing the Mean and the Median

```{r, echo = FALSE, fig.margin=TRUE, fig.height=4, fig.cap="Normal Distribution"}
 N <- 100000
 x <- rnorm(n=N, mean=100, sd=10)
 hist(x, 
 xlim=c(min(x),max(x)), probability=T, nclass=max(x)-min(x)+1, 
   col='lightblue', xlab=' ', ylab=' ', axes=F,
   main='Symmetic Distribution')
#lines(density(x,bw=1), col='red', lwd=3)
abline(v = mean(x),                     # Add line for mean
       col = "green",
       lwd = 3)
abline(v = median(x),                    # Add line for median
       col = "red",
       lwd = 3)
```

In distributions that are symmetric, we expect that the mean and median will 
be extremely close if not equal. In the example shown in Figure 1, the mean
and median are 100.03 and 100.04 respectively. The values are so close that 
their lines overlap in Figure 1.

```{marginfigure}
**Outlier**:
An unusually small or unusually large data value.
```

This is not always the case, however. Consider the example offered by Figure 2.
We can see the mean and median pulling apart. In fact, the median (the red line)
is closer to the typical value in this distribution. The mean is pulled toward
the extreme values in this distribution because the mean as a measurement is 
sensitive to **outliers**, unusual values in a distibution. The median, on the
other hand, is relatively insensitive to outliers. Often the median is the 
measure of central tendency reported with economic or business data because
measurements related to money are often skewed like the data shown in Figure 2.

```{r, echo = FALSE, fig.margin=TRUE, fig.height=4, fig.cap="Skewed Distribution"}
 N <- 100000
 #x <- rnbinom(N, 10, .3)
 x<-  rexp(N,0.15)
 hist(x, 
 xlim=c(min(x),max(x)), probability=T, nclass=max(x)-min(x)+1, 
   col='lightblue', xlab=' ', ylab=' ', axes=F,
   main='Positive Skew')
#lines(density(x,bw=1), col='red', lwd=3)
abline(v = mean(x),                     # Add line for mean
       col = "green",
       lwd = 3)
abline(v = median(x),                    # Add line for median
       col = "red",
       lwd = 3)
```

# Mode

```{marginfigure}
**Mode**:
The value that occurs with the greatest frequency.
```

In our running problem, **46, 54, 42, 46, 32**, the mode is 46 as it is the
only value that occurs twice in the data set.

# Percentiles

A **percentile** provides information about how the data are spread over the 
interval from the smallest to the largest value. A percentile divides the 
data into two parts: the data less than the percentile and the data greater
than the percentile. For example, the 25th percentile divides the
data such that 25% of the data are less than the 25th percentile and 75% of the 
data are greater than the 25th percentile.

```{marginfigure}
**Location of the Pth Percentile**: 
\Large
\begin{equation} 
    L{p} = \frac{p}{100} (n+1)
\end{equation}
where $p$ is the percentile and $n$ is the number of observations.
```

To calculate the 25th percentile, we begin by ordering our data: \
\
**46, 54, 42, 46, 32** --> **32, 42, 46, 46, 54**\
\
After sorting the data, we use formula 3 to calculate the position or rank
of the 25th percentile, $L_{25}$. In this case, we substitute 25 for $p$ and 5 for $n$.\
\
$L_{25} = \frac{25}{100} (5+1) = 1.5$\
\
So we now have the rank or position of the 25th percentile, $L_{25}$. 
We need to convert this to an actual value. 
The rank number 1.5 tells us the following:\
\
1. **1**.5 tells us that the 25th percentile is after position 1 (value 32).\
2. 1**.5** tells us that the 25th percentile is .5 (half) of the way between positions 1 and 2.\
\
To convert this information to a value, we perform the following calculation:\
\
25th percentile = 32 + (42-32)\*0.5 = `r 32 + (42-32)*0.5`
\


# Measures of Location and Scales of Measurement

In our previous sessions, we introduced scales of measurement to help us
understand when certain techniques are appropriate given the properties of our
data. Scales of Measurement is particularly useful when thinking about 
reasonable uses of measures of central tendency. Table 1 summarizes the 
measures of location most appropriate for different scales of measurement.

Scale of Measurement|Measure of Location
--------------------|-------------------
Nominal             | Mode
Ordinal             | Median, Mode
Interval            | Arithmetic Mean, Median, Mode
Ratio               | Geometric Mean, Arithmetic Mean, Median, Mode
