---
title: "Descriptive Statistics: Tabular and Graphical"
runningheader: "Descriptive Statistics 1" # only for pdf output
subtitle: "Probability and Statistics" # only for html output
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library('plyr')
library(epiDisplay)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

Statistical work begins with accurately counting. **Frequency distributions**
and **histograms** are common and effective ways to summarize the basic data we
collect.

# Frequency Distributions

```{marginfigure}
**Frequency Distribution**: 
A frequency distribution is a tabular summary of data showing the number 
(frequency) of observations in each of several nonoverlapping categories or
classes. 
```

Frequency distributions are useful for summarizing categorical and quantitative
data. Consider the following example related to categorical data. In this 
example, we collected data on 20 soda purchases.

```{r, echo=FALSE, fig.fullwidth=TRUE}
soda <- c("Coke","Pepsi","Sprite","Fanta","Dr. Pepper",
          "Coke","Pepsi","Coke","Pepsi","Fanta",
          "Coke","Coke","Coke","Pepsi","Dr. Pepper",
          "Coke","Pepsi", "Coke", "Fanta", "Coke")
print(soda)
print(count(soda))
```

```{marginfigure}
**Relative Frequency**: 
\Large
\begin{equation} 
    \frac{frequency}{n}
\end{equation}
where frequency is the class frequency and n is the number of observations 
in the sample data set.
```

When we count the number of times people purchased each type of soda, we 
arrive at the frequency distribution displayed above. Figure 1 demonstrates
how we can present this data graphically with a bar chart. In some cases, we
want to see the **relative frequency** of each class. We can display the 
data as a **relative frequency distribution** as well. The relative frequency
of a class is the frequency of the class divided by the number of observations
in the data set. For example, the relative frequency of Coke is 0.45 or 
9 divided by 20 (n). 

```{r, echo=FALSE, fig.fullwidth=TRUE}
rf <- table(soda)/length(soda)
print(count(rf)[,1:2])
```

```{r, echo=FALSE, fig.margin=TRUE, fig.fullwidth=FALSE, fig.cap="Bar Chart", fig.height=5}
barplot(table(soda),width=c(5,5,5,5,5), cex.names=.5, ylim=c(0,10))
usr <- par("usr")
par(usr=c(usr[1:2], 0, 10))
axis(2,at=seq(0,10,1))
```

Frequency distributions can also be used for quantitative variables. In these
cases, we must be a bit more thoughtful about how we construct the overlapping
classes. In particular, we must pay close attention to:

1. The number of nonoverlapping classes we want;
2. The width of each class;
3. The class limits.

In the following example, we collected 20 student exam grades.

```{r, echo=FALSE, fig.fullwidth=TRUE}
grades <- c(80,85,75,81,90,
            70,79,84,74,80,
            95,75,83,88,78,
            55,65,98,73,89)

print(grades)
```

In order to make a meaningful frequency distribution, we need to select the 
number of classes that we want. Five classes of width 10 makes good sense
since we are grouping exam grades. The table below shows the class boundaries
which coincide with the familiar grade breakdowns from A to F. The table
includes the frequency distribution and the relative frequency distribution as
in our previous example. 

```{r, echo=FALSE, fig.fullwidth=TRUE}
br = seq(49,99,by=10)

ranges = paste(head(br,-1)+1, br[-1], sep=" - ")
freq   = hist(grades, breaks=br, include.lowest=TRUE, plot=FALSE)

df <- data.frame(range = ranges, frequency = freq$counts)

df[,3] <- df[,2]/length(grades)

df[,4] <- cumsum(df[,2])

df[,5] <- df[,4]/length(grades)

names(df) <- c("Class", "Frequency", "Relative Freq.", "Cumulative Freq.", "Cum. Relative Freq.")

print(df)
```

The table also includes two new variations. The 
**cumulative frequency distribution** shows the number of data items with
less than or equal to the upper class limit of each class. 
The **cumulative relative frequency distribution** shows the proportion of the
data items with values less than or equal to the upper limit of each class.
Figure 2 demonstrates how we can visualize the frequency distribution of a quantitative 
variable with a **histogram**.

```{r, echo=FALSE, fig.margin=TRUE, fig.fullwidth=FALSE, fig.cap="Histogram", fig.height=5}
hist(grades,breaks=br)
```

