---
title: "Descriptive Statistics: Tabular and Graphical"
runningheader: "Descriptive Statistics 1" # only for pdf output
subtitle: "Probability and Statistics" # only for html output
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

We see **data** every day, and yet we often lack the language to 
describe, the concepts to understand, and the tools to interpret those data.
This course aims to provide you with the language, concepts, and tools you 
need to comprehend data and harness their power.

```{marginfigure}
**Data**: 
The facts and figures collected, analyzed, and summarized for presentation
and interpretation. 
```

# Basic Vocabulary

```{marginfigure}
**Element**: 
The entity on which data are collected. 
```

```{marginfigure}
**Variable**: 
A characteristic of interest for an element. 
```

```{marginfigure}
**Observation**: 
The set of measurements collected for a particular element. 
```

```{marginfigure}
**Data Set**: 
All collected data for a particular study. 
```

In Figure 1, we see the common vocabulary related to data in action.
**Elements** are the entities on which data are collected. In Figure 1, 
the element is the company, represented by its name. **Variables** are
characteristics of interest for the elements. For example, for each
element (company), variables for Stock Exchange, Annual Sales, and
Earnings per share were collected. An **observation** is the set of 
measurements collected for a particular element. In this case, the row marked
Observation contains the set of measurements (variables) related to the element,
EnergySouth. Finally, **data set** refers to all the data collected for the
particular study. In this case, the data set comprises 5 observations
measuring 3 variables for each element.

# Types of Data and Scale of Measurement

```{marginfigure}
**Categorical Data**: 
Labels or names used to identify an attribute of each element. 
```

```{marginfigure}
**Quantitative Data**: 
Numeric values that indicate how much or how many of something of interest. 
```

Variables use different kinds of numbers to identify or measure attributes.
Consider two different types of numbers related to athletes. In many sports,
athletes wear jerseys to identify them from other players on the field. Jersey
numbers are labels that identify particular players. Compare a jersey number
to an athlete's performance statistics or height measurements. Where the 
jersey number is a label, an athlete's height is a proper quantitative 
measurement. Variables such as jersey number are referred to as 
**categorical variables** ; those like height are referred to 
as **quantitative variables**.


While the distinction between categorical and quantitative variables is 
useful, data analysts and statisticians often require more precision
when considering how to responsibly use variables.
The **scale of measurement** determines the amount of information contained in 
the data and indicates the most appropriate data aggregation and 
statistical analyses. Table 1 summarizes the scales of measurement, their 
properties, and examples.

```{marginfigure}
**Nominal Scale**: 
The scale of measurement for a variable when the data are labels or names
used to identify an attribute of an element. 
```

```{marginfigure}
**Ordinal Scale**: 
The scale of measurement for a variable if the data exhibit the properties
of nominal data and the order or rank of the data is meaningful. 
```

```{marginfigure}
**Interval Scale**: 
The scale of measurement for a variable if the data demonstrate the properties
of ordinal data and the interval between values is expressed in terms of a
fixed unit of measure. 
```

```{marginfigure}
**Ratio Scale**: 
The scale of measurement for a variable if the data demonstrate all the 
properties of interval data and the ratio of two values is meaningful. 
```

```{r, fig.fullwidth=TRUE, echo=FALSE}
df <- data.frame(c("Nominal","Ordinal","Interval","Ratio"),
                 c("Y","Y","Y","Y"),
                 c("N","Y","Y","Y"),
                 c("N","N","Y","Y"),
                 c("N","N","N","Y"),
                 c("Gender","Rankings","Temperature","Ratio"))

names(df) <- c("Scale","Named","Ordered",
                "Proportionate Intervals", "Absolute Zero", "Examples")

knitr::kable(
  df, caption = 'Scales of Measurement'
)
```

Figure 2 summarizes the types of data and scales of measurement in a taxonomy.
The purpose of thinking through the scales of measurement of variables is 
to ensure that apply sound techniques to data with those properties. Consider
the example of athletes again. While taking the average height of basketball or 
volleyball players makes sense, taking an average of the jersey numbers of 
those players makes no sense at all.



