---
title: "Trade Secrets"
runningheader: "Review of DTSA Year 1" # only for pdf output
subtitle: "Review of DTSA Year 1" # only for html output
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(dplyr)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

This paper constructs decision tree models based on the data published here:
[DTSA Data Sets](https://www.dtsalitigation.com/2018/04/30/datasets/)

The data sets support the publication found here:
[DTSA Year One](https://scholarlycommons.law.wlu.edu/cgi/viewcontent.cgi?article=1523&context=wlufac)

# Read Case Data

For a full description of the coding for this case file see:
[DTSA Data Sets](https://www.dtsalitigation.com/2018/04/30/datasets/)

```{r}
cases <- read.csv(file = 'cases.csv')
# str(cases)
drop <- c("plaintiff","defendant","date","month","court_n","judge",
          "fedlaw1","fedlaw2","fedlaw3","fedlawother","statelaw1","statelaw2",
          "statelaw2","statelaw3","statelaw4","statelaw5","statelaw6",
          "statelaw7","statelaw8","statelaw9","statelaw10","statelawother",
          "secret1","secret2","secret3","secret4","secret5","secret6",
          "secret7","secret8","secretother","relation1",
          "relation2","relation3","relation4","notes",
          "link_complaint")
df_cases = cases[,!(names(cases) %in% drop)]
df_cases[,"docket"] <- trimws(df_cases[,"docket"])
df_cases[,"court"] <- trimws(df_cases[,"court"])
#str(df_cases)

```

# Read Relief Data

For a full description of the coding for this relief file see:
[DTSA Data Sets](https://www.dtsalitigation.com/2018/04/30/datasets/)

```{r}
relief <- read.csv(file = 'relief.csv')
drop <- c("plaintiff","defendant","date_dtsa","district_n",
          "coder","notes")
df_relief = relief[,!(names(relief) %in% drop)]
df_relief[,"docket"] <- trimws(df_relief[,"docket"])
colnames(df_relief)[colnames(df_relief) == 'district'] <- 'court'
df_relief[,"court"] <- trimws(df_relief[,"court"])
#str(df_relief)
```

# Join Data Sets on Docket and Court

Two coding errors in the relief file produced two missing records.
Docket numbers 3:17-CV-00004 and 2:17-CV-00075 had incorrectly coded
Courts/Disctricts. (DND and DNJ do not appear in the appendix of the 
coding document.) These are typos for NDD and NJD. We have cleansed data.


```{r}
df <- merge(x=df_cases,y=df_relief,by=c("docket","court"))
#str(df)
```

Checking for duplicates and missing records.
Anti-joins to find missing or extra cases.

```{r}
df_anti <- df_relief %>% anti_join(df, by=c("docket","court"))
#head(df_anti)

df_anti2 <- df_cases %>% anti_join(df, by=c("docket","court"))
#head(df_anti2)
```

# Temporary Restraining Orders (TRO)

```{r}
drop <- c("pi","exparte","docket", "court","nos","jur_fq","jur_div","jur_supp","remove","country")
df_tro = df[,!(names(df) %in% drop)]

#Convert all variables to factors
col_names <- names(df_tro)
df_tro[,col_names] <- lapply(df_tro[,col_names] , factor)

#For this task, tro=1 should be removed. They did not seek this remedy.
df_tro <- df_tro[df_tro$tro != 1, ]
#str(df_tro)

fit <- rpart(tro~., data = df_tro, method = 'class', cp = 0.0001, maxdepth=3,
             parms = list(split = 'information'),  minsplit = 2, minbucket = 1)
#summary(fit)
fit
#rpart.plot(fit, box.palette = "white")
fancyRpartPlot(fit, caption = NULL)

```

# Preliminary Injunction (PI)

```{r}
drop <- c("tro","exparte","docket", "court","nos","jur_fq","jur_div","jur_supp","remove","country")
df_pi = df[,!(names(df) %in% drop)]

#Convert all variables to factors
col_names <- names(df_pi)
df_pi[,col_names] <- lapply(df_pi[,col_names] , factor)

#For this task, pi=1 should be removed. They did not seek this remedy.
df_pi <- df_pi[df_pi$pi != 1, ]
#str(df_tro)

fit <- rpart(pi~., data = df_pi, method = 'class', cp = 0.0001, maxdepth=3,
             parms = list(split = 'information'),  minsplit = 2, minbucket = 1)
#summary(fit)
fit
#rpart.plot(fit, box.palette = "white")
fancyRpartPlot(fit, caption = NULL)

```

# Ex Parte

```{r}
drop <- c("tro","pi","docket", "court","nos","jur_fq","jur_div","jur_supp","remove","country")
df_ex = df[,!(names(df) %in% drop)]

#Convert all variables to factors
col_names <- names(df_ex)
df_ex[,col_names] <- lapply(df_ex[,col_names] , factor)

#For this task, pi=1 should be removed. They did not seek this remedy.
df_ex <- df_ex[df_ex$exparte != 1, ]
#str(df_tro)

fit <- rpart(exparte~., data = df_ex, method = 'class', cp = 0.0001, maxdepth=3,
             parms = list(split = 'information'),  minsplit = 2, minbucket = 1)
#summary(fit)
fit
#rpart.plot(fit, box.palette = "white")
fancyRpartPlot(fit, caption = NULL)

```
