---
title: "Worked Example: Standard Error of the Mean"
runningheader: "Standard Error of the Mean" # only for pdf output
subtitle: "An implementation in R Markdown" # only for html output
author: "Lombardi"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Standard Error of the Mean

```{r fig-margin1, fig.margin = TRUE, fig.cap = "Normal Distribution with mean, 100, standard deviation, 10.", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE, echo=FALSE}
#define population mean and standard deviation
population_mean <- 100
population_sd <- 10

#define upper and lower bound
lower_bound <- population_mean - population_sd
upper_bound <- population_mean + population_sd

#Create a sequence of 1000 x values based on population mean and standard deviation
x <- seq(-4, 4, length = 1000) * population_sd + population_mean

#create a vector of values that shows the height of the probability distribution
#for each value in x
y <- dnorm(x, population_mean, population_sd)

#plot normal distribution with customized x-axis labels
plot(x,y, type = "l", lwd = 2, axes = FALSE, xlab = "", ylab = "")
sd_axis_bounds = 5
axis_bounds <- seq(-sd_axis_bounds * population_sd + population_mean,
                    sd_axis_bounds * population_sd + population_mean,
                    by = population_sd)
axis(side = 1, at = axis_bounds, pos = 0)
abline(v=population_mean, col="blue")
abline(v=population_mean+population_sd, col="gray")
abline(v=population_mean-population_sd, col="gray")
```

In many cases, we cannot collect data related to every entity in a population.
Practical constraints like effort and cost often require us to sample the 
population to obtain an estimate of the measurement we seek.\
When estimating values like the population mean, we often need a sense of 
the precision of our estimates. As it turns out, if we repeatedly
sample our population, the sample means will 
cluster around the population mean. The standard error of the mean, 
$\sigma_{\bar x}$, measures the dispersion of the the sample means.\
\
```{marginfigure}
We compute a *sample statistic* to estimate the value of a *population
parameter*. For example, the sample mean $\bar{x}$ is the
point estimator of the population mean, $\mu$. In other words, $\bar{x}$ 
provides an estimate of $\mu$.
```
\
The *central limit theorem* helps us understand how the sample means are
distributed around the population mean.\  
\
```{marginfigure}
**Central Limit Theorem**: 
In selecting random samples of size n from a population, the sampling distribution of
$\bar{x}$ can be approximated by a *normal distribution* as the sample size becomes large.
```
\
This worked example outlines two different techniques for calculating the 
standard error of the mean. 

# Standard Error of the Mean (Infinite Population)

**Problem:** Suppose a random sample of size 50 (n=50) is selected from a 
population with 
$\sigma = 10$. Find the value of $\sigma_{\bar x}$ assuming that the 
population size is infinite.\
\
*Step 1:* Formula 1 provides the standard error of the mean, 
$\sigma_{\bar x}$. Apply the forumula to our problem's specific values.\  
\
```{marginfigure}
\Large
\begin{equation} 
    \sigma_{\bar x} = \frac{\sigma}{\sqrt{n}}
\end{equation}
Use this formula whenever the population is infinite or the sample size
is less than or equal to 5% of the population size.
```  
\
After substituting the values, the formula should look like:\
\
$\sigma_{\bar x} = \frac{10}{\sqrt{50}}$\
\
*Step 2:* Perform the calculation\
\
$\sigma_{\bar{x}}$ = `r 10/sqrt(50)`  

# Standard Error of the Mean (Finite Population)

**Problem:** Suppose a random sample of size 50 (n=50) is selected from a 
population with 
$\sigma = 10$. Find the value of $\sigma_{\bar x}$ assuming that the 
population size (N=500).\
\
*Step 1:* Formula 2 provides the standard error of the mean $\sigma_{\bar x}$
with finite population correction factor. 
Apply the forumula to our problem's specific values.\  
\
```{marginfigure}
\Large
\begin{equation} 
    \sigma_{\bar x} = \sqrt{\frac{N-n}{N-1}} \left( \frac{\sigma}{\sqrt{n}} \right)
\end{equation}
Use this formula whenever the population is finite and the sample size
is greater than 5% of the population size.
```  
\
After substituting the values, the formula should look like:\
\
$\sigma_{\bar x} = \sqrt{\frac{500-50}{500-1}} \left( \frac{10}{\sqrt{50}} \right)$\
\
*Step 2:* Perform the calculation\
\
$\sigma_{\bar{x}}$ = `r sqrt((500-50)/(500-1))*(10/sqrt(50))`  

# Sampling Distribution

The **sampling distribution of $\bar{x}$** is the probability distribution of 
all possible values of the sample mean $\bar{x}$.

```{r fig-margin2, fig.margin = TRUE, fig.cap = "Sampling Distribution", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE, echo=FALSE}
#make this example reproducible
set.seed(0)

#define number of samples
n = 10000

#create empty vector of length n
sample_means = rep(NA, n)

#fill empty vector with means
for(i in 1:n){
  sample_means[i] = mean(rnorm(n=10, mean=100, sd=10))
}

#create histogram to visualize the sampling distribution
hist(sample_means, main = "", xlab = "Sample Means (n=10)", col = "steelblue")
```

The central limit theorem suggests that a large, properly 
drawn sample will resemble the population from which it is drawn
(Wheelan, 2013). The sampling distribution of sample means from repeatedly 
drawn samples will cluster around the population mean even if the distribution
of the population is skewed. In other words, the distribution of sample 
means is normal.\
Consider Figures 2, 3, and 4.\ Figure 2 shows the sampling distributon 
for 10000 independently drawn samples of
n=10 for $\mu = 100$ and $\sigma = 10$ from a normal distribution.\
$\bar{x}=$ `r mean(sample_means)` -  $\sigma_{\bar{x}}=$ `r 10/sqrt(10)`\

```{r fig-margin3, fig.margin = TRUE, fig.cap = "Sampling Distribution", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE, echo=FALSE}
#make this example reproducible
set.seed(0)

#define number of samples
n = 10000

#create empty vector of length n
sample_means = rep(NA, n)

#fill empty vector with means
for(i in 1:n){
  sample_means[i] = mean(rnorm(n=50, mean=100, sd=10))
}

#create histogram to visualize the sampling distribution
hist(sample_means, main = "", xlab = "Sample Means (n=50)", col = "steelblue")
```

Figure 3 shows the sampling distribution for 10000 independently drawn samples
of n=50 for $\mu = 100$ and $\sigma = 10$ from a normal distribution.\
$\bar{x}=$ `r mean(sample_means)` -  $\sigma_{\bar{x}}=$ `r 10/sqrt(50)`\

```{r fig-margin4, fig.margin = TRUE, fig.cap = "Sampling Distribution", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE, echo=FALSE}
#make this example reproducible
set.seed(0)

#define number of samples
n = 10000

#create empty vector of length n
sample_means = rep(NA, n)

#fill empty vector with means
for(i in 1:n){
  sample_means[i] = mean(rnorm(n=150, mean=100, sd=10))
}

#create histogram to visualize the sampling distribution
hist(sample_means, main = "", xlab = "Sample Means (n=150)", col = "steelblue")
```

Figure 4 shows the sampling distribution for 10000 independently drawn samples
of n=150 for $\mu = 100$ and $\sigma = 10$ from a normal distribution.\
$\bar{x}=$ `r mean(sample_means)`  -  $\sigma_{\bar{x}}=$ `r 10/sqrt(150)`\
As the sample size (n) increases, the sample means group more closely to
their expected value, the population mean.

